Reproducible Research: Peer Assessment 1
=========================================

### Loading and preprocessing the data

Before running the code we assume that the working directory is set up correctly and that "activity.zip" file containing data has been extracted, so that "activity.csv" file is now located in main directory. Otherwise, we should execute the code that was commented in the first lines of a code chunk below. 

We start with loading data from the csv file and converting "date" column 
to a "Date" format.


```{r readData}
# setwd("~/Documents/edu/datasciencecoursera/5 Reproducible research/hw/RepData_PeerAssessment1")
#unzip("activity.zip")

activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
activity[,2] <- as.Date(activity[,2])
```

### What is mean total number of steps taken per day?

```{r histogram: total steps per day, fig.path = "figure/"}
stepsPerDay <- xtabs(steps ~ date, data = activity)
m <- mean(stepsPerDay)
med <- median(stepsPerDay)

# png(file = "figure/histogram1.png")
hist(stepsPerDay, breaks = 20)
# dev.off()

```

The mean of total number of steps taken per day is `r m` and median is `r med`.


## What is the average daily activity pattern?


```{r average daily activity pattern, fig.path = "figure/"}
library(reshape2)
m <- melt(activity, id = "interval", variable.name = "steps", 
          na.rm = TRUE)
d <- dcast(m,  interval ~ "steps", mean)

# png(file = "figure/averActivity.png" )
plot(d[,1], d[,2], type = "l", xlab = "5-min interval", 
     ylab = "Average number of steps taken", 
     main = "Average daily activity pattern")
# dev.off()

idx <- which.max(d[,2])
```
Maximum number of steps (`r max(d[,2]) ` steps) were made in this 5-minute interval: `r d[idx,1]`. 


## Imputing missing values


```{r Imputing missing values, fig.path = "figure/"}
# sum(is.na(activity))

#get time intervals for each row with missing number of steps
time <- activity[is.na(activity),3]
#vector of values used to fill the NAs
intervalAver <- c() 

for (i in 1:length(time)) {
        intervalAver <- c(intervalAver, d[d[,1] == time[i],"steps"])
        }

#fill NAs with the average number of steps for given 5 minute period
completeData <- activity      #create new data set
completeData[is.na(completeData),1] <- intervalAver

stepsPerDayCompl <- xtabs(steps ~ date, data = completeData)
#head(stepsPerDayCompl,20)
hist(stepsPerDayCompl, breaks = 20)

# 
# png(file="figure/filled_missing_values.png")
# hist(stepsPerDayCompl, breaks = 20)
# dev.off()

# mean(stepsPerDayCompl) == mean(stepsPerDay)
# 
# as.vector(stepsPerDay)
# as.vector(stepsPerDayCompl)
rbind(summary(as.vector(stepsPerDay)),
      summary(as.vector(stepsPerDayCompl)))
# median(stepsPerDayCompl) == median(stepsPerDay)

# boxplot(as.vector(stepsPerDay), as.vector(stepsPerDayCompl))
# sum(as.vector(stepsPerDay))
# sum(as.vector(stepsPerDayCompl))

```

There are `r sum(is.na(activity))` rows with missing values in this data set. 
After filling the missing values new data set has different median (`r median(stepsPerDayCompl)` vs. previously `r median(stepsPerDay)`) and much higher mean (`r mean(stepsPerDayCompl)` vs. previously `r mean(stepsPerDay)`).


## Are there differences in activity patterns between weekdays and weekends?


```{r weekdays vs. weekend, fig.path = "figure/"}

wd <- factor(weekdays(completeData[,2]), levels = c("poniedziałek","wtorek", "środa", "czwartek", "piątek","sobota","niedziela"))
weekendDays <- levels(wd)[6:7]

#create a logical vector - is this a weekend day?
isWeekend <- wd %in% weekendDays

#create factor variable "day": "weekend" or "weekday" 
completeData[isWeekend,"day"] <- "weekend"
completeData[!isWeekend,"day"] <- "weekday"
completeData[,"day"] <- as.factor(completeData[,"day"]) 
        
# summary(completeData)
# summary(as.factor(completeData$day))


#mean steps by "day"
library(dplyr)
# summary(completeData)
# 
# select(completeData, -date)
# filter(completeData, day == "weekend")
# mutate(completeData, )
# summarise(completeData, meansteps = mean(steps))


bytime <- group_by(completeData, interval, day)
# summarise(bytime, mean(steps))

meanStepsWeekend <- completeData %.%
        group_by(day, interval) %.%
        summarise(meansteps = mean(steps)) %.%
        filter(day == "weekend")

meanStepsWeekday <- completeData %.%
        group_by(day, interval) %.%
        summarise(meansteps = mean(steps)) %.%
        filter(day == "weekday")

# png(file = "figure/weekend_vs_weekdays.png")
par(mfcol = c(2,1))
plot(meanStepsWeekday[,2], meanStepsWeekday[,3], type = "l", 
     main = "Weekadays: average number of steps during the day",
     xlab = "Time", ylab = "Steps per 5 min. intervals")
plot(meanStepsWeekend[,2], meanStepsWeekend[,3], type = "l", 
     main = "Weekend",
     xlab = "Time", ylab = "Steps per 5 min. intervals")
par(mfcol = c(1,1))
# dev.off()

```
